# (PART) Data Import and Export {-}

From hereonin we're going to be focusing on using R, rather than the history, community and everything else. One approach to teaching a programming language is to step through the different rules in how the language is constructed, what the different keywords are, any strange quirks it has, and all the rest - and this can be really helpful for people who are *already* programmers.

That isn't guaranteed, though, and so the rest of this book is divided up into individual, practical applications of the language, combined with explanations of what's going on in each piece so that you can learn the structure of the language through doing.

So, going forward, you'll see something that looks like the following; I'll introduce a concept, or a step in analysing data, and talk about why it's important. And then, after a colon, because I flagrantly abuse both the colon and semicolon:


```{r, eval=FALSE}
There'll be a chunk of code, in a shaded box just like this one, which will achieve the step I was talking about
```

After that will be an explanation of that chunk of code, and what each bit is doing and how everything in it works. If you're the sort of person who instead wants to jump into the nitty-gritty of the language, I recommend Hadley Wickham's *[Advanced R Programming](http://adv-r.had.co.nz/Data-structures.html)* (specifically the data structures section) to get a handle for the general workings of R, and then coming back here to learn the applications.

# Reading data from files

Obviously the first step in data analysis is getting the data, so let's start there! Datasets come in lots of different formats, so we'll cover the ones you most commonly see in a business context, namely:

1. Comma-separated and tab-separated values files (CSV/TSV);
2. Microsoft Excel spreadsheets;
3. MySQL and similar databases.

## CSV & TSV

CSVs and TSVs are the most common file format for data, and the good news there is that as a consequence, R has a whole host of tools for reading them in. Let's open by loading one of them: the `readr` add-on I had you install earlier:

```{r, eval=FALSE}
library(readr)
```

So, what's this code doing? First, `library`; that's a function, which is a self-contained chunk of code that's packaged and made available so you can use it as many times as you want without having to write the code out entirely each time. `readr` is the input provided to `library`. So it works out as `do_this_thing(to_this_thing)`.

The `library` function loads *packages* - the add-ons I was talking about earlier - so you can refer to the code they contain when you're working. In this case it's loading `readr`, a package specifically designed for reading in and writing out a wide range of file formats, including CSVs and TSVs.

So we have our file-reading code. Now let's read in the file, with:

```{r, eval=FALSE}
medical_data <- read_csv("https://www.ironholds.org/misc/medical_data.csv")
```

Let's break down what's going on in that code, going from right to left. First we have a piece of text - a URL, in fact - in quote marks. The quote marks identify the text as a `string`; something that R should treat just as text, not as code to be executed.

That URL is wrapped in parentheses, preceded by `read_csv`: as we've seen from loading the `readr` package, that's a function call. So we have a call to run `read_csv`, with the URL as an argument. What it's doing is reading in a CSV, from that URL!

Then we shift left and run into `<-`. That's R's *assignment operator*; it assigns the results of whatever code is on the right of the operator, to a name on the left of it (in this case, `medical_data`). So the code, in plain English, is "read in the CSV at this URL and assign the contents of it to the name `medical_data`". The result (the output of code, tied to a name) is an *object*.

We can take a look at the structure of an object just by typing its name into the console. Let's do that now:

```{r, eval=FALSE}
medical_data
```

As you can see, it's tabular, just like the CSV - columns of values, rows of observations. What you're looking at is a "data frame", the most common type of object in R: it's designed to hold tabular data, since tabular data is what most statisticians and data scientists rely on.

We can get a more detailed look at it with the `str` function (which means "structure" and does exactly what it says on the tin):

```{r, eval=FALSE}
str(medical_data)
```

## Excel

## Other types

# Reading data from databases

# Exporting data
