# (PART) Introduction to R {-}

# Background

## A brief history of splines

## The breadth of R

## The R community

# Getting set up

## Installation
To get started with R, you're going to need...well, R!

If you're on Windows you can get R [here](https://cran.r-project.org/bin/windows/base/); if you're on a mac, you can download it [here](https://cran.r-project.org/bin/macosx/). For Linux users (specifically Ubuntu users, or people on other, Debian-based distributions) open the terminal and type:

```{r, eval=FALSE}
sudo apt-get -y install r-base
```

This requires the password to your computer's administrator account; if you don't have it, you'll need to get the system administrator to install R.

Once you've got R, you'll also need RStudio - an Integrated Development Environment (IDE) for R that we'll be covering shortly. RStudio can be obtained [here](https://www.rstudio.com/products/rstudio/download/).

After that, you'll need a few R packages (more on those later) that are used in this tutorial. Open RStudio, now that you've installed it, and go to the window in the bottom left marked "console". Type:

```{r, eval=FALSE}
install.packages("readr",
                 "dplyr",
                 "tidyr",
                 "shinydashboard",
                 "magrittr")
```

## Using RStudio

You should now have RStudio; great! That's going to be your primary 

# The language

## Basic operations in R

## Objects

## Functions

## Packages
